A Constructive Metatheory of Bidirectional Type Synthesis for Simple Types

### 2023-06-13 tex/sec3-defs.tex
[ tex ] Split BiSig.tex into smaller files

\documentclass[acmsmall,screen]{acmart}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[british]{babel}

\usepackage{bm,braket,mathtools,mathpartir,bussproofs}
\EnableBpAbbreviations
\usepackage[all,cmtip]{xy}

\usepackage{subcaption,xfrac,xspace,xcolor,xspace}
\usepackage{hyperref}
\usepackage[capitalise,noabbrev]{cleveref}
\usepackage[inline]{enumitem} % for environment enumerate*
\setlist[enumerate]{mode=unboxed}

\setlength {\marginparwidth }{2cm}
\usepackage[color=yellow,textsize=small]{todonotes}
\newcommand{\LT}[1]{\todo[author=LT,inline,color=green!40,caption={}]{{#1}}}
\newcommand{\Josh}[1]{\todo[author=Josh,inline,color=green!40,caption={}]{{#1}}}

\usepackage{subfiles}

\usepackage{microtype}

\citestyle{acmauthoryear}

\AtEndPreamble{%
\theoremstyle{acmdefinition}
\newtheorem{remark}[theorem]{Remark}}

\input{type-notation} %% for type theory notation copied from HoTT Book
\input{macros}

\begin{document}

\author{Liang-Ting Chen}
\email{liang.ting.chen.tw@gmail.com}
\orcid{0000-0002-3250-1331}
\author{Hsiang-Shang Ko}
\orcid{0000-0002-2439-1048}
\email{joshko@iis.sinica.edu.tw}

\affiliation{%
  \institution{Academia Sinica}
  \streetaddress{128 Academia Road}
  \city{Taipei}
  \country{Taiwan}
  \postcode{115}
}

\title{A Constructive Metatheory of Bidirectional Type Synthesis for Simple Types}

\begin{abstract}
  This paper proposes a formal theory of bidirectional typing.
\end{abstract}

\maketitle

\section{Introduction}\label{sec:intro}

\begin{enumerate}
  \item Type checking as a transition from (parsed) abstract syntax trees to well-typed terms.
  \item Bidirectional typing is the only solution that works that scales well.
  \item Not yet a general theory but by now a design principle.
    No `type checker generator' as each type checker is still designed and implemented on a case-by-case basis, but can be derived from a syntax-directed type theory `easily'.
  \item The existing work of bidirectional typing does not take account of missing annotations carefully enough. 
  \item A meta-problem: existing work is based on (or at least left unspecified) classical logic.
    One has to state soundness of completeness of their type checking algorithm in a less straightforward way.
  \item A review of what has been done towards this line of research.
  \item Finally, our responses (contributions of this paper).
\end{enumerate}

\subsection{Related work}

\citep{Xie2018}
\subsubsection{Language formalization and its frameworks}
\cite{Wadler2022}
PoplMark Challenge~\citep{Aydemir2005}

\cite{Cimini2020,Cimini2022}

\citep{Ahrens2018,Fiore2022,Gheri2020,Ahrens2022}
\cite{Allais2021}

\subsubsection{Theories of abstract syntax with variable binding}
\cite{Fiore1999,Hirschowitz2010,Ahrens2018,Fiore2022,Ahrens2021,Arkor2020,Hirschowitz2022}
\cite{Fiore2013,Hamana2011,Hamana2022}

\subsubsection{Type checker generation}
\cite{Gast2004,Grewe2015,Pacak2020,Cimini2020}

\subsubsection{Bidirectional typing}

\cite{Pierce2000}
\cite{Dunfield2021}

\subsection{Plan and Contributions of this paper}

Our assumptions and contributions:
\begin{enumerate}
  \item Assumption: Syntax-directed typing rules
  \item Practical: bidirectionalisation over annotatability, decidability over soundness and completeness.
  \item Generic: Type checking once and for all bidirectional type systems for simple types.
  \item Correct and Constructive: Fully formalised in Martin-Löf type theory (with Axiom K).
    LEM is not assumed, so a decidable statement has both computational and non-trivial logical readings.
\end{enumerate}

Plan of this paper:
\begin{enumerate}
  \item Key ideas (\Cref{sec:key-ideas})
  \item Definitions for simple type theories and for bidirectional type systems (signature erasure and mode annotation) (\Cref{sec:defs})
  \item Soundness, completeness, bidirectionalisation, and annotatability (\Cref{sec:annotatability})
    \Josh{Erasure from bidirectional typing derivation to raw terms with a mode}
  \item Bidirectional type inference (\Cref{sec:type-synthesis})
  \item Formalisation with further examples (\Cref{sec:formalisation})
  \item Future work (\Cref{sec:future})
\end{enumerate}
  \LT{better title?}

\section{Core Ideas with Bidirectional Simply Typed \texorpdfstring{$\lambda$}{lambda}-Calculus} \label{sec:key-ideas}
\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Gamma \vdash \isTerm{t} \Rightarrow A}
    \quad\text{and}\quad \boxed{\Gamma \vdash \isTerm{t} \Leftarrow A}
    \\
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Gamma \vdash \isTerm{t\subsum} \Leftarrow A}\;(\text{sub})
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Gamma \vdash \isTerm{u} \Leftarrow A}{\Gamma \vdash \isTerm{t\;u} \Rightarrow B}\;(\text{app})
    \and
    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \Leftarrow B}{\Gamma \vdash \isTerm{\lam{x}t} \Leftarrow A \bto B}\;(\text{abs})
  \end{mathpar}
  \caption{Bidirectional simply typed $\lambda$-calculus $\Lambda^{\Leftrightarrow}_{\bto}$}
  \label{fig:bi-stlc}
\end{figure}
\begin{remark}
  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
\end{remark}

@@

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}}
    \quad\text{where $\vdash_{\Sigma, \erase{\Omega}} \isTerm{t}$}
    \\
    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}^\isDir{\syn}}\;(\text{var})
    \and
    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}^\isDir{\chk}}{\vdash_{\Sigma, \Omega} (\isTerm{t \annote A})^\isDir{\syn}}\;(\text{anno})
    \and
    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{\syn}}{\vdash_{\Sigma, \Omega} {\isTerm{t\subsum}}^\isDir{\chk}}\;(\text{ sub})
  \end{mathpar}
  \begin{mathpar}
    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1}^\isDir{d_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}^\isDir{d_n}}
    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}^\isDir{d}}\;(\text{op})
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
  \end{mathpar}
  \caption{Raw terms in some mode $d$ for a bidirectional type system $(\Sigma, \Omega)$}
\end{figure}

@@

\section{Soundness, Completeness, and Bidirectionalisation} \label{sec:annotatability}

\subsection{Soundness and Completeness}
\begin{theorem}[Soundness]\label{thm:term-soundness}

\end{theorem}

\begin{theorem}[Completeness]\label{thm:term-completeness}

\end{theorem}

\subsection{Bidirectionalisation}
\begin{proposition}[Decidability for Bidirectionalisation] \label{thm:bidirectionalisation}
  For every raw term $t$ and a mode $d$, it is decidable that $t$ is in mode $d$, i.e.\ $\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}$ is derivable.
\end{proposition}

\subsection{Completeness versus Annotatability}

\section{Bidirectional Type Synthesis and Checking} \label{sec:type-synthesis}

\subsection{Mode Correctness}

\begin{theorem}[Uniqueness of Synthesised Types]\label{thm:unique-syn}
  Let $(\Sigma, \Omega)$ be a bidirectional type system where $\Omega$ is mode-correct.
  Then, for every raw term $t$ in synthesising mode and two derivations of 
  \[
    \Gamma \vdash_{\Sigma, \Omega} t \Rightarrow A
    \quad\text{and}\quad
    \Gamma \vdash_{\Sigma, \Omega} t \Rightarrow B
  \]
  for some types $A$ and $B$ respectively, the synthesised types must be equal, i.e.\ $A = B$.
\end{theorem}

\subsection{Trichotomy of Raw Terms by Type Synthesis}

\begin{theorem}[Decidability of Bidirectional Typing Synthesis] \label{thm:bidirectional-type-synthesis}
  For every type signature $\Sigma$ and bidirectional binding signature $\Omega$ for $\Sigma$ that is mode-correct, the following two statements hold.

  \begin{enumerate}
    \item For every context $\Gamma$ and raw term $\isTerm{t}$ in synthesising mode, it is decidable that there is a type~$A$ and a derivation of
      \[
        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A.
      \]
    \item For every context $\Gamma$, raw term $\isTerm{t}$ in checking mode, and type $A$, it is decidable that there is a derivation of
      \[
        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A.
      \]
  \end{enumerate}
\end{theorem}

\begin{corollary}[Trichotomy of Raw Terms]
  Let $\Sigma$ be a type signature and $\Omega$ a bidirectional binding signature.
  Then, for every context $\Gamma$ and raw term $t$, exactly one of the following statements holds:
  \begin{enumerate}
    \item $t$ is synthesising and there exists a typing derivation of\, $\Gamma \vdash_{\Sigma, \Omega} t : A$ for some type $A$.
    \item $t$ is synthesising but there is no typing derivation of\, $\Gamma \vdash_{\Sigma, \Omega} t : A$ for any type $A$.
    \item $t$ is not synthesising.
  \end{enumerate}
\end{corollary}
\begin{proof}
  Combine  \Cref{thm:term-soundness,thm:term-completeness,thm:bidirectionalisation} with \Cref{thm:bidirectional-type-synthesis}.

\end{proof}

\subsection{Decidability of Bidirectional Type Synthesis}

\begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis}}]
  By induction on raw terms $t$ in some mode and for the $(\text{op})$ rule on the list of arguments.
  Moreover, if the type $A$ of $t$ is synthesised and checked against $B$ but $A \neq B$, then \Cref{thm:unique-syn} is applied to derive a contradiction.

  \begin{enumerate}
    \item 
        \LT{explain all cases but operation here}
%    \item For every mode-correct operation $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$, a context $\Gamma$, and a raw term
%      \[
%        \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)}, 
%      \]
%      with a mode $d$, it is decidable that there exists a substitution $\rho$ from $\Xi$ to $\emptyset$ and a typing derivation of
%      \[
%        \Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)} : \simsub{A}{\rho}^\isDir{d}.
%      \]
    \item For every set $\Xi$ of type variables, list of
      \[
        \left(\Delta_i, A_i, d_i\right) : \Cxt_{\Sigma}(\Xi) \times \Type_{\Sigma}(\Xi) \times \Mode, 
      \]
      \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$, context $\Gamma$, and raw terms $t_i$'s in mode $d_i$,
      \begin{enumerate}
        \item either there is a minimal extension $\ext{\rho}$ of $\rho$ such that all     of following judgements are derivable
          \[
            \Gamma, \simsub{\Delta_i}{\ext{\rho}} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\ext{\rho}}^{d_i}
          \]

        \item or there is no extension $\ext{\rho}$ such that all $\Gamma, \simsub{\Delta_i}{\ext{\rho}} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\ext{\rho}}^{d_i}$ has a typing derivation. 
      \end{enumerate}
  \end{enumerate}
\end{proof}

\section{Formalisation and More Examples} \label{sec:formalisation}
\LT{how to type Agda code without lhs2tex?}

\begin{enumerate}
  \item Bidirectional binding signature, functor, and terms (extrinsic typing, raw terms, raw terms with a mode, i.e.\ pre-typed terms).
  \item type synthesis and checking.
  \item examples. 
\end{enumerate}

\section{Future Work} \label{sec:future}

\Josh{Getting rid of the assumption of being syntax-directed (the triangular picture)}

\paragraph{Polymorphic Algebraic Theories}
\cite{Fiore2013,Hamana2011,Hamana2022}

\begin{acks}
We thank Nathanael Arkor for the useful conversation and to thank the Programming Languages and Formal Methods Laboratory at Academia Sinica for the opportunity of sharing earlier ideas.

The work is supported by the Ministry of Science and Technology of Taiwan under grant MOST 109-2222-E-001-002-MY3.
\end{acks}

\bibliographystyle{ACM-Reference-Format}

\IfFileExists{reference.bib}{%
  \bibliography{reference}
}{\bibliography{ref}}

\end{document}

>>>>

%! TEX root = BiSig.tex

@@

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}}
    \quad\text{where $\vdash_{\Sigma, \erase{\Omega}} \isTerm{t}$}
    \\
    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}^\isDir{\syn}}\;(\text{var})
    \and
    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}^\isDir{\chk}}{\vdash_{\Sigma, \Omega} (\isTerm{t \annote A})^\isDir{\syn}}\;(\text{anno})
    \and
    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{\syn}}{\vdash_{\Sigma, \Omega} {\isTerm{t\subsum}}^\isDir{\chk}}\;(\text{sub})
  \end{mathpar}
  \begin{mathpar}
    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1}^\isDir{d_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}^\isDir{d_n}}
    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}^\isDir{d}}\;(\text{op})
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
  \end{mathpar}
  \caption{Raw terms in some mode $d$ for a bidirectional type system $(\Sigma, \Omega)$}
\end{figure}

@@
